---
title: "Getting to Know TEDS-A"
author: "Amber Camp"
date: "`r Sys.Date()`"
format: html
editor: visual
toc: true
---

## Before We Start

Get to know your data well enough to figure out *which* questions it can actually answer. Every dataset has a personality!

A note on this notebook: chunks that start with `#| eval: false` are **yours** to complete. Fill in the blanks (`___`), then delete the `#| eval: false` line so the chunk runs when you render.

## Load Libraries

If you don't have the below packages, run the relevant lines here:

```{r}
# install.packages("tidyverse")
# install.packages("janitor")
# install.packages("naniar")
```

Load libraries (all of them).

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor) # clean_names() and tabyl()
library(naniar) # tools for exploring and visualizing missing data
options(scipen = 99) # turns off scientific notation
```

## TEDS-A Data

Treatment Episode Data Set-Admissions (TEDS-A), 2023, from the Substance Abuse and Mental Health Services Administration (SAMHSA). The codebook can be found [here](https://www.samhsa.gov/data/system/files/media-puf-file/teds-a-2023-ds0001-info-codebook_v1.pdf).

Make sure `tedsa_puf_2023.rdata` is saved in the `data` folder of your project.

### Read the Codebook First (Seriously!)

Before you touch the data, open the codebook and answer the following. Write your answers right here in the notebook. These answers are part of your data audit.

1.  Read the introduction and the "Universe" section. **What is one row in this dataset?** Be specific.

    **Your answer:**

2.  Read "Coverage" and "Data Limitations." **Name at least three kinds of people or treatment settings that will *not* show up in this data.**

    **Your answer:**

3.  Look at Appendix A. Some variables come from the "Minimum data set" and some from the "Supplemental data set." **What's the difference, and why might it matter for missing data?**

    **Your answer:**

4.  Find the "State Exclusions" table. **Which states are excluded in 2023?**

    **Your answer:**

5.  Use Ctrl+F (Cmd+F) to search the codebook for the word "mental." **Which variables come up, and what do they actually measure? What does that tell you about the kind of treatment this dataset describes?**

    **Your answer:**

## Loading an `.rdata` File

So far we've mostly read in data with functions like `read_csv()`, where *we* choose the name of the dataframe. An `.rdata` file works differently: `load()` drops objects into your environment with whatever name they were saved under. So we need to ask R what it just loaded.

```{r}
loaded_name <- load("data/tedsa_puf_2023.rdata")
loaded_name # what's the name of the object we just loaded?
```

Now copy that object into a dataframe with a name we like, and clean up the column names. I suggest `teds_a` for now.

```{r}
teds_a <- get(loaded_name) %>%
  clean_names() # makes all column names lowercase with underscores

# remove the original copy so we aren't holding two giant dataframes in memory
if (loaded_name != "teds_a") rm(list = loaded_name)
```

### First Look

```{r}
dim(teds_a)
glimpse(teds_a)
```

How many rows and columns are there? What *type* does R think most of these columns are? Maybe try `skimr`.

**Your answer:**

### One Row Is Not One Person

`caseid` is unique for every row. Let's confirm.

```{r}
n_distinct(teds_a$caseid) == nrow(teds_a)
```

Go back to the "Data Limitations" section of the codebook. If `caseid` is unique for every row, what is it identifying? Can one person show up more than once?

**Your answer:**

## Numbers That Aren't Numbers

R thinks most of these columns are numbers. Let's see what happens when we treat them that way.

```{r}
class(teds_a$age)
summary(teds_a$age)
```

According to R, the "mean age" is somewhere around 7. Is anyone in this dataset 7 years old? Look up `AGE` in the codebook.

**Your answer:**

Here's the big idea: **most of the columns in TEDS are categories wearing number costumes.** A `5` in `age` doesn't mean five years, it means a *bin* of ages. A `15` in `stfips` doesn't mean Hawaiʻi is "more" than Alaska (`2`). If you calculate a mean, a standard deviation, or a correlation on these codes, R will happily give you an answer, and it will be meaningless.

## Zooming In on Hawaiʻi

How many admissions does each state have?

```{r}
state_counts <- teds_a %>%
  tabyl(stfips) %>%
  arrange(desc(n))

state_counts
```

Filter down to Hawaiʻi.

```{r}
hawaii_fips <- 15 # confirm this in the STFIPS section of the codebook. where does this code come from?

teds_hi <- teds_a %>%
  filter(stfips == hawaii_fips)

nrow(teds_hi)
```

The national file is big. If you don't need national comparisons right now, remove it to free up memory. (You can always reload it later.)

```{r}
rm(teds_a)
invisible(gc()) # "garbage collection" - asks R to actually release the memory
```

How many admissions are in Hawaiʻi? What might that number mean once you start splitting the data into smaller groups (by age, by substance, by service type...)?

**Your answer:**

## Missing Data (Where Is It Hiding?)

What data is missing? Examine with `naniar`.

```{r}
miss_var_summary(teds_hi)
```

No missing data at all? That seems... too good to be true. Let's look closer at one variable.

```{r}
teds_hi %>%
  tabyl(employ)
```

What is `-9`? Look up `EMPLOY` in the codebook and find the label for `-9`.

**Your answer:**

TEDS doesn't use `NA`. It codes missing values as `-9`, and R has no way of knowing that `-9` means "missing" unless we tell it. Let's tell it, for every column at once.

```{r}
teds_hi <- teds_hi %>%
  mutate(across(everything(), ~ na_if(.x, -9))) # across every column, turn -9 into NA
```

Inspect once again.

```{r}
miss_var_summary(teds_hi)

gg_miss_var(teds_hi)

vis_miss(teds_hi) # our Hawaiʻi subset is small enough to visualize all of it
```

Which variables are mostly or entirely missing for Hawaiʻi? Compare a couple of them to the national percentages listed in the codebook. Is Hawaiʻi different?

**Your answer:**

### Not All Missing Is the Same

Now that everything is `NA`, it's easy to forget *why* something is missing. Try this:

```{r}
teds_hi %>%
  tabyl(sex, preg)
```

Read the `PREG` entry in the codebook. Why are all of those values missing for one group?

**Your answer:**

Now read the descriptions for `DETCRIM` and `DETNLF`. When would those be missing on purpose?

**Your answer:**

There are (at least) three different kinds of "missing" hiding in `NA`:

-   **Not applicable** - the question doesn't apply to this person

-   **Not collected** - Hawaiʻi doesn't report this variable at all

-   **Unknown** - it applied, and it wasn't recorded

These mean very different things for your analysis. How you handle them is a researcher choice, and you should be able to explain yours.

## From Codes to Labels

Numbers are hard to read and easy to misuse. Turn codes into labeled categories using `factor()`. We'll keep the original column and create a new labeled one right next to it, so we can check our work.

### Example 1: Service Setting

`services` is a *nominal* variable: categories with no natural order. Copy the labels from the `SERVICES` section of the codebook.

```{r}
teds_hi <- teds_hi %>%
  mutate(services_lbl = factor(services,
                               levels = 1:8,
                               labels = c("Detox, 24-hour, hospital inpatient",
                                          "Detox, 24-hour, free-standing residential",
                                          "Rehab/residential, hospital (non-detox)",
                                          "Rehab/residential, short term (30 days or fewer)",
                                          "Rehab/residential, long term (more than 30 days)",
                                          "Ambulatory, intensive outpatient",
                                          "Ambulatory, non-intensive outpatient",
                                          "Ambulatory, detoxification")))
```

**Always check your recode.** [Every code should line up with exactly one label]{.underline}.

```{r}
teds_hi %>%
  tabyl(services, services_lbl)
```

Since there's no natural order, sorting by frequency makes a readable plot.

```{r}
ggplot(teds_hi, aes(x = fct_infreq(services_lbl))) +
  geom_bar() +
  coord_flip() +
  labs(x = "Service setting at admission",
       y = "Number of admissions",
       title = "Hawaiʻi admissions by service setting",
       subtitle = "TEDS-A 2023 (SAMHSA)")
```

### Example 2: Age

`age` is *ordinal*: the categories have a meaningful order. Here, sorting by frequency would scramble that order, so we don't.

```{r}
teds_hi <- teds_hi %>%
  mutate(age_lbl = factor(age,
                          levels = 1:12,
                          labels = c("12-14", "15-17", "18-20", "21-24",
                                     "25-29", "30-34", "35-39", "40-44",
                                     "45-49", "50-54", "55-64", "65+")))

teds_hi %>%
  tabyl(age, age_lbl)

ggplot(teds_hi, aes(x = age_lbl)) +
  geom_bar() +
  labs(x = "Age at admission",
       y = "Number of admissions",
       title = "Hawaiʻi admissions by age group",
       subtitle = "TEDS-A 2023 (SAMHSA)")
```

What do you notice about the bin widths? What does that mean if you want to talk about "younger" vs "older" clients?

**Your answer:**

### What About the NAs?

By default, `factor()` leaves `NA` as `NA`, and `ggplot` will show it as its own bar. If you want missing values to have a visible, named category in tables and plots, you can do this:

```{r}
teds_hi %>%
  mutate(age_lbl = fct_na_value_to_level(age_lbl, level = "Missing")) %>%
  tabyl(age_lbl)
```

Whether you keep, label, or drop missing values is (you guessed it) a researcher choice.

## String Gotchas

Once your codes become labels, you're working with text. Text has to match *exactly*.

```{r}
"Ambulatory, detoxification" == "Ambulatory, detoxification"
"Ambulatory, detoxification" == "Ambulatory, detoxification " # one sneaky space at the end
"Ambulatory, detoxification" == "ambulatory, detoxification"  # capitalization counts
```

If you ever `filter()` for a label and get zero rows back, check for spaces and capitalization before you panic. `str_squish()` removes extra whitespace, and `str_to_lower()` standardizes capitalization.

### Collapsing Categories with `str_detect()`

Sometimes eight categories is too many, especially with a small sample. `str_detect()` checks whether a piece of text appears inside a string, which makes it handy for grouping labels. The `^` means "starts with."

```{r}
teds_hi <- teds_hi %>%
  mutate(setting_group = case_when(
    str_detect(services_lbl, "^Detox") ~ "Detox",
    str_detect(services_lbl, "^Rehab") ~ "Residential",
    str_detect(services_lbl, "^Ambulatory") ~ "Outpatient"
  ))

teds_hi %>%
  tabyl(services_lbl, setting_group)
```

Where did "Ambulatory, detoxification" end up? Is that where you would put it?

Try adding a line that catches "detox" *anywhere* in the label (no `^`, and watch the capitalization!). `case_when()` checks conditions **in order** and stops at the first match, so does it matter whether your new line goes first or last? Which grouping makes more sense, and why?

**Your answer:**

### Codes as Text: Leading Zeros

If you ever join TEDS to another dataset (Census data, for example), you may find that the other dataset stores state codes as text with a leading zero, like `"02"` for Alaska. A number can't hold a leading zero, so you'd need to convert:

```{r}
tibble(stfips = c(2, 6, 15)) %>%
  mutate(state_fips = sprintf("%02d", stfips)) # "pad this number to 2 digits with zeros"
```

## Your Turn: Recode Your Variables

Choose the variables your research question needs. For each one, look it up in the codebook, decide whether it's nominal or ordinal, and recode it. Then **check your recode**.

```{r}
#| eval: false
teds_hi <- teds_hi %>%
  mutate(___ = factor(___,
                      levels = ___,
                      labels = c(___)))

teds_hi %>%
  tabyl(___, ___)
```

Copy that chunk as many times as you need.

## Practice

**1. What percentage of Hawaiʻi admissions list alcohol as the primary substance?**

Hint: `tabyl()` gives you percents. `adorn_pct_formatting()` makes them readable.

```{r}
#| eval: false
teds_hi %>%
  tabyl(___) %>%
  adorn_pct_formatting()
```

**2. For each service setting, what share of admissions are missing employment status?**

Hint: `group_by()`, `summarize()`, and `mean(is.na(___))`. (Why does `mean()` of TRUE/FALSE give you a proportion?)

```{r}
#| eval: false
teds_hi %>%
  group_by(___) %>%
  summarize(n = n(),
            prop_missing = mean(is.na(___)))
```

**3. Write your own.** Write a plain-language question that uses at least one of *your* variables, then translate it into code.

**Your question:**

```{r}
#| eval: false

```

## Quick EDA

Make at least two plots of the variables you recoded. Choose your ordering thoughtfully (nominal vs. ordinal), and label your axes like a person who respects their audience.

```{r}
#| eval: false
ggplot(teds_hi, aes(x = ___)) +
  geom_bar() +
  labs(x = "___",
       y = "Number of admissions",
       title = "___",
       subtitle = "TEDS-A 2023 (SAMHSA)")
```

With a smaller sample, some bars may be very short. Before you interpret a difference between two groups, check the actual counts behind it.

## Save Your Work

You've put a lot of work into this dataframe! Save it so you can read it straight in next time. We'll use `.rds`, which keeps your factor levels and labels intact (a `.csv` would not).

```{r}
write_rds(teds_hi, here("data/teds_hi_clean.rds"))

# next time, read it back in with:
# teds_hi <- read_rds(here("data/teds_hi_clean.rds"))
```

## Before Next Class

As a group, write a short data audit summary covering:

-   What one row represents, and who is *not* in this data

-   How many Hawaiʻi admissions there are, and what that means for your analysis

-   The variables you selected, why each one connects to your research question, and how much of each is missing for Hawaiʻi

-   **Can TEDS-A answer your research question as written?** If not, what would a dataset need to contain in order to answer it?

Then revisit your research question against our five criteria (answerable, specific, non-obvious, consequential, scoped). Which criterion did the audit change the most?
